

---
title: 'Reproducible Research - Project 1: Activity Montioring Data Analysis'
author: "SrChip15"
date: "June 10, 2016"
output: html_document
---


```{r setoptions, echo=TRUE}
```


Let's download the dataset from the given URL and load the dataset into our WD
and after that remove the temp file that we no longer need
```{r part 1(a)}
# download the relevant dataset from an URL source and load it into workspace
URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
temp <- tempfile()
download.file(URL, temp)
org.df <- read.csv(unz(temp, "activity.csv"), stringsAsFactors = FALSE)
unlink(temp)
```


Now that the dataset has been loaded, let's get some statistics on the dataset
to get an overview of the dataset
```{r exploratory}
# explore dataset
dim(org.df)
summary(org.df)
str(org.df)
```

Right off the bat, we observe that the "date" variable is interpreted as a
character datatype. Let's go ahead and convert it into the appropriate date format
```{r part 1(b)}
# convert date from char to date
org.df$date <- as.Date(org.df$date)
str(org.df)
```
The "str" call now confirms that the said variable is in the appropriate date format


Now, let's aggregate the dataset to get the total number of steps taken in a day
```{r part 2(a)}
#  aggregate number of steps on date
agg_date.df <- setNames(aggregate(org.df$steps ~ org.df$date, data = org.df, FUN = sum), c("Date","Steps"))
head(agg_date.df)
```
As we can observe from the "head" call, now we ONLY see UNIQUE dates, assuring 
that the steps have been aggregated over unique dates


For the first part of the assignment, let's graph a histogram of the total number
of steps taken each day and also report the mean and median of the total number
of steps taken per day
```{r part 2(b)}
# load library and make a histogram (gg)
library(ggplot2)
ggplot(agg_date.df, aes(Steps)) +
        geom_histogram(binwidth = 2000, fill = "blue") +
        labs(title = "Histogram of Daily Steps", x = "Steps per Day", y = "Frequency")
```

```{r part 2(c)}
# mean and median
mean_org <- mean(agg_date.df$Steps)
median_org <- median(agg_date.df$Steps)
print(mean_org)
print(median_org)
```
The mean and the median differ by 1.19 steps, not significantly different from each other
implying that the datapoints are failry equally distributed


Moving on to the next question in the assignment - Average Daily Activity Pattern.
For this purpose, we would be averaging the total number of steps taken in a day
over the 5-minute intervals for all the days present in the dataset.

Further, we report the interval with the maximum number of steps
```{r part 3(b)}
# averaging the number of steps over 5-minute intervals 
agg_int.df <- setNames(aggregate(org.df$steps ~ org.df$interval, data = org.df, FUN = mean), c("Interval","Steps"))

# report the interval with the highest mean number of steps 
max_int_avg <- agg_int.df[which.max(agg_int.df$Steps),]
print(max_int_avg)
```
It is clear now that during the months of October and November of 2012, 835th 5-minute
interval is the time of the day that has the highest average number of steps in a day


Let's visualize the same by way of a time series plot
```{r part 3(a)}
# time series plot of average steps across intervals
ggplot(agg_int.df, aes(x = Interval, y = Steps)) +
        geom_line(color = "blue") +
        labs(title = "Time Series of Daily Average Steps over Intervals", x = "Intervals", y = "Number of Steps")
```


Missing values! Like any real dataset, the current dataset is also riddled with 
missing values. So, let's analyze the number of missing value cases and also the 
corresponding variables
```{r part 4(a)}
# number of missing values
missing_val <- sapply(org.df, function(x) sum(is.na(x)))
print(missing_val)
```
Approximately, missing value cases constitute 13.11% of the dataset and this is
entirely from the "steps" variable in the dataset

This is a fairly significant proportion of the dataset. So, a simple mean value
imputation is enforced in the dataset
```{r part 4(b & c)}
# copy original dataset
imp.df <- org.df

# impute missing values with the mean
imp.df$steps[is.na(imp.df$steps)] <- mean(imp.df$steps, na.rm = TRUE)
str(imp.df)
```


Now let's again aggregate the number of steps over unique dates and report the 
mean and median of the same

```{r part 4 (d)(i)}
# aggregate the imputed data set to get the total number of steps taken daily
agg_date_imp.df <- setNames(aggregate(imp.df$steps ~ imp.df$date, data = imp.df, FUN = sum), c("Date","Steps"))

# mean and median of imputed dataset
mean_imp <- mean(agg_date_imp.df$Steps)
median_imp <- median(agg_date_imp.df$Steps)
print(mean_imp)
print(median_imp)
```
Now the mean and the median have the same value. Without imputation, the mean and 
the median differed by 1.19 steps. So, in this case, since the original data (with 
missing values) is equally distributed, there is none to negligibel impact on the
dataset. Factually, we are able to see that the mean and median are now equal


Visualizing the total number of steps after imputing for missing values
```{r part 4 (d)(ii)}
# histogram of total number of steps taken each day (with imputed values)
ggplot(agg_date_imp.df, aes(Steps)) +
        geom_histogram(binwidth = 2000, fill = "blue") +
        labs(title = "Histogram of Daily Steps (w/ imputed val)", x = "Steps per Day", y = "Frequency")
```


Before we wrap up, let's analyze the dataset for activity distributions from a
weekday Vs. weekend standpoint. As a result, we would need to engineer a new feature, 
which I would title as "DayType". Basically, this would be a 2-level factor variable, 
namely - Weekday or Weekend.
```{r part 5 (a)}
# add DayType (weekend/weekday) feature into the imputed dataset
weekend <- c("Saturday","Sunday")
imp.df$DayType <- as.factor(ifelse(weekdays(imp.df$date) %in% weekend, "Weekend", "Weekday"))
head(imp.df)
str(imp.df)
```


Okay, now let's average the number of steps taken in the imputed dataset over 
intervals for all the days
```{r part 5 (b)(i)}
# aggregate data points over intervals
agg_int_imp.df <- setNames(aggregate(imp.df$steps ~ imp.df$interval + imp.df$DayType, data = imp.df, FUN = mean), c("Interval","DayType", "Steps"))
head(agg_int_imp.df)
tail(agg_int_imp.df)
```


Finally, let's visualize the mean number of steps over different intervals in a 
weekday Vs. weekend perspective to observe a pattern (if any)
```{r part 5 (b)(ii)}
# time series panel plot for Interval Mean Steps - Weekday Vs. Weekend
ggplot(agg_int_imp.df, aes(x = Interval, y = Steps, color = DayType)) +
        geom_line() +
        facet_wrap(~DayType, ncol = 1, nrow = 2) +
        labs(title = "Interval Mean Steps - Weekday Vs. Weekend", x = "Interval", y = "Mean Steps")
```


From the sharp peaks in the weekdays, we can conclude that during the weekdays
there are time periods when there is peak activity and the rest of the day is fairly
low activity. The peak activity falls between 1pm and 3pm, presumably users heading 
out for lunch or taking a break from work. While, over the weekends, it looks like
the users are consistently active throughout the day and there are no drastic peaks
that signal peak activity times.

